---
title: "Population-Weighted Maternal Health Coverage Analysis"
subtitle: "Comparing On-Track vs Off-Track Countries for SDG 3.1"
author: "_"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
  pdf_document:
    toc: true
  word_document:
    toc: true
---

**Positions Applied For:**
- Household Survey Data Analyst Consultant - Req. #581656
- Microdata Harmonization Consultant – Req. #581699

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)

# Load required libraries
library(dplyr)
library(readr)
library(ggplot2)
library(knitr)
library(scales)
library(viridis)
library(tidyr)
library(stringr)
```

## Executive Summary

This analysis compares population-weighted maternal health coverage between countries that are **on-track** versus **off-track** for achieving SDG 3.1 (reducing maternal mortality). Using 2022 projected births as weights, we examine coverage gaps in **Antenatal Care (ANC4)** and **Skilled Birth Attendance (SBA)**.

## Data and Methodology

The analysis combines three key datasets:
- **UNICEF maternal health indicators** (ANC4 and SBA coverage, 2018-2022)
- **UN World Population Prospects 2022** (projected births for 2022 as weights)
- **SDG 3.1 country classifications** (on-track vs off-track status)

Population-weighted averages ensure that countries with larger populations have proportional influence on the results, providing a more accurate global picture.

```{r load-data}
# Load the analysis results
weighted_coverage <- read_csv("../04_output/population_weighted_coverage.csv")
merged_df <- read_csv("../04_output/merged_data.csv")

# Display data summary
kable(weighted_coverage, 
      caption = "Population-Weighted Maternal Health Coverage by Track Status",
      digits = 1)
```

## Key Findings

```{r coverage-gaps}
# Calculate coverage gaps
on_track <- weighted_coverage[weighted_coverage$track_status == "on-track", ]
off_track <- weighted_coverage[weighted_coverage$track_status == "off-track", ]

anc4_gap <- round(on_track$anc4_weighted_avg - off_track$anc4_weighted_avg, 1)
sba_gap <- round(on_track$sba_weighted_avg - off_track$sba_weighted_avg, 1)
```

- **ANC4 Coverage Gap**: `r anc4_gap` percentage points (On-track: `r on_track$anc4_weighted_avg`% vs Off-track: `r off_track$anc4_weighted_avg`%)
- **SBA Coverage Gap**: `r sba_gap` percentage points (On-track: `r on_track$sba_weighted_avg`% vs Off-track: `r off_track$sba_weighted_avg`%)

## Visualization: Coverage Comparison

```{r coverage-visualization, fig.cap="Population-Weighted Maternal Health Coverage: On-Track vs Off-Track Countries"}
# Prepare data for visualization
plot_data <- weighted_coverage %>%
  select(track_status, anc4_weighted_avg, sba_weighted_avg) %>%
  pivot_longer(cols = c(anc4_weighted_avg, sba_weighted_avg),
               names_to = "indicator", 
               values_to = "coverage") %>%
  mutate(
    indicator = case_when(
      indicator == "anc4_weighted_avg" ~ "ANC4 (4+ Antenatal Visits)",
      indicator == "sba_weighted_avg" ~ "SBA (Skilled Birth Attendance)"
    ),
    track_status = stringr::str_to_title(track_status)
  )

# Create the main comparison chart
p1 <- ggplot(plot_data, aes(x = indicator, y = coverage, fill = track_status)) +
  geom_col(position = "dodge", width = 0.7, alpha = 0.8) +
  geom_text(aes(label = paste0(coverage, "%")), 
            position = position_dodge(width = 0.7), 
            vjust = -0.5, size = 4, fontface = "bold") +
  scale_fill_manual(values = c("Off-Track" = "#e74c3c", "On-Track" = "#27ae60"),
                    name = "Track Status") +
  scale_y_continuous(limits = c(0, 100), 
                     labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Population-Weighted Maternal Health Coverage Comparison",
    subtitle = "Coverage gaps between On-Track and Off-Track countries for SDG 3.1",
    x = "Maternal Health Indicator",
    y = "Coverage (%)",
    caption = "Note: Weighted by projected births for 2022. Data sources: UNICEF, UN WPP 2022"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray60"),
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 12, face = "bold"),
    panel.grid.minor = element_blank(),
    plot.caption = element_text(size = 9, color = "gray50")
  )

print(p1)
```

```{r country-distribution, fig.cap="Country and Birth Distribution by Track Status"}
# Additional visualization showing country and birth distribution
dist_data <- weighted_coverage %>%
  select(track_status, total_countries, total_births_millions) %>%
  pivot_longer(cols = c(total_countries, total_births_millions),
               names_to = "metric", values_to = "value") %>%
  mutate(
    metric = case_when(
      metric == "total_countries" ~ "Number of Countries",
      metric == "total_births_millions" ~ "Total Births (Millions)"
    ),
    track_status = stringr::str_to_title(track_status)
  )

p2 <- ggplot(dist_data, aes(x = track_status, y = value, fill = track_status)) +
  geom_col(alpha = 0.8, width = 0.6) +
  geom_text(aes(label = ifelse(grepl("Countries", metric), 
                              as.character(value), 
                              paste0(value, "M"))), 
            vjust = -0.5, fontface = "bold") +
  facet_wrap(~metric, scales = "free_y") +
  scale_fill_manual(values = c("Off-Track" = "#e74c3c", "On-Track" = "#27ae60")) +
  labs(
    title = "Distribution of Countries and Births by Track Status",
    x = "Track Status", y = "Value"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.position = "none",
    strip.text = element_text(face = "bold")
  )

print(p2)
```

## Detailed Results Table

```{r detailed-table}
# Create a comprehensive results table
results_table <- weighted_coverage %>%
  select(track_status, anc4_weighted_avg, sba_weighted_avg, 
         total_countries, total_births_millions) %>%
  mutate(track_status = stringr::str_to_title(track_status))

kable(results_table,
      col.names = c("Track Status", "ANC4 Weighted (%)", "SBA Weighted (%)", 
                    "Total Countries", "Total Births (M)"),
      caption = "Population-Weighted Coverage Analysis by Track Status",
      digits = 1)
```

## Interpretation and Key Insights

The analysis reveals significant disparities in maternal health coverage between countries on-track and off-track for achieving SDG 3.1:

**Coverage Gaps**: Off-track countries demonstrate substantially lower population-weighted coverage for both indicators. The ANC4 gap of `r anc4_gap` percentage points and SBA gap of `r sba_gap` percentage points highlight the maternal health challenges in countries struggling with high maternal mortality rates.

**Population Impact**: The population-weighted approach reveals the true global burden, as it accounts for the number of women and births affected. Countries with larger populations have proportional influence on the results, providing a more accurate picture of worldwide maternal health access.

**Policy Implications**: The coverage gaps underscore the need for targeted interventions in off-track countries, particularly focusing on improving antenatal care access and ensuring skilled attendance at birth.

### Caveats and Assumptions

1. **Data Availability**: Coverage estimates represent the most recent data available for each country (2018-2022), which may not reflect current conditions in all cases.

2. **Population Weights**: Analysis uses UN 2022 projected births as weights, assuming these projections accurately represent the target population for maternal health services.

3. **Country Classifications**: SDG 3.1 track status is based on progress toward maternal mortality reduction targets, which may not perfectly correlate with service coverage patterns.

4. **Missing Data**: Countries without complete data for track status, birth projections, or coverage indicators were excluded from the analysis, potentially affecting representativeness.

5. **Temporal Alignment**: Coverage data spans 2018-2022 while using 2022 birth projections as weights, assuming coverage patterns remain relatively stable within this timeframe.

## Conclusions

This analysis demonstrates clear maternal health coverage disparities between countries on different trajectories for SDG 3.1 achievement. The population-weighted approach provides valuable insights into the global burden of inadequate maternal health access, highlighting the urgent need for targeted interventions in off-track countries to ensure equitable progress toward universal health coverage.

---

**Positions Applied For:**
- Household Survey Data Analyst Consultant - Req. #581656
- Microdata Harmonization Consultant – Req. #581699

---

*Report generated on `r Sys.Date()` using R Markdown*